Glucose ingestion before and after resistance training does not augment ribosome biogenesis in healthy moderately trained young adults

### 2023-02-03 ribose-paper.RMD
updated manuscript

Regular resistance exercise training (RT) increases muscle mass and muscle strength, due to the adaptations stimulated by a mechanical strain. Unfortunately, responses to RT are not uniform across the population. Studies have shown RT-induced muscle growth to vary widely between individuals, with as much as 10-15% showing considerably impaired growth (ALVAREZ18, MANN14, THALACKER13). This may be explained by genetics, epigenetics, and/or an unfavourable internal physiological milieu (KILDER). Recently, tweaking training modalities and -protocols have been shown to induce different responses in different populations (HAMMARSTROM20), however, little is known regarding the optimisation of RT for individual phenotypes. Generally, meta-analyses favour a moderate volume of RT over low volume RT regarding muscle growth and -strength (KRIEGER09, RHEA03, SCHOENFELD17), substantiated by recent RCTs (KILDER). Although Hammarström and colleagues (20) found that a moderate volume led to more prominent gains in muscle mass and -strength, approximately 50% of the participants did not experience true benefits of moderate volume compared to low, with a blunted muscle growth (and strength?). Due to the variations observed in muscle growth responses to standardised RT protocols (HAMMARSTROM20; SCHOENFELD17; STEC16?), it is apparent that increasing RT volume alone does not convert low-responders to high-responders. Therefore, other means than RT per se may be necessary to circumvent this discrepancy.

Indeed, ingesting nutrients such as protein and creatine supplements effectively optimise RT (CERMAK12, LANHERS15, LANHERS17). However, it remains unknown if other nutritional adjuvants such as glucose can increase the efficacy of RT. This is surprising, as glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major supplier of energy to cells via ATP synthesis (TANAKA18). Energy availability is important to regulate central processes of muscle growth, specifically ribosome biogenesis, a key factor for increased muscle protein synthesis (MPS) (TANAKA18, FIGUEIREDO19). Interestingly, variations in RT-induced responses have been linked to ribosome biogenesis, where participants responding poorly to RT also showed a blunted ribosome biogenesis (HAMMARSTROM20, STEC16). A blunted ribosome biogenesis negates responses to RT by failure to produce novel ribosomes, thus the inability to increase the skeletal muscle cell’s translational capacity (VONWALDEN19, FIGUEIREDO19). Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into the regulation of ribosome biogenesis.

Ribosome biogenesis is suggested to be regulated by the mammalian target of rapamycin complex 1 (mTORC1) network (KUSNADI15, TANAKA18). Stimulated by growth factors, hormones, mechanical loading and nutrients, mTORC1 governs energy distribution by turning on/off energy-consuming and energy-generating pathways, depending on cellular energy levels (HOPPE09). Ribosome biogenesis consumes up to 80% of the cell’s energy and materials (SCHMIDT04) and is presumably switched on/off based on cellular energy levels (VONWALDEN19). Upon stimulation by mTORC1, ribosome biogenesis is initiated with the assembly of the preinitiation complex (PIC) (VONWALDEN19, FIGUEIREDO19). The PIC consists of the RNA polymerase I complex, the SL-1 complex and the TIF-1A/RRN3 protein (FIGUEIREDO19, VONWALDEN19). Further, the upstream binding factor (UBF) has a vital role in the initiation, binding the PIC to the rDNA promoter, while c-Myc serves as a transcription factor (FIGUEIREDO19, VONWALDEN19). Upon initiation, Pol I transcribed the 47S pre-rRNA, ultimately processed into the 18S-, 5.8S and 28S rRNA (MULLINEUX12, VONWALDEN19). The fourth rRNA, 5S, is transcribed by Pol III, while mRNAs for the ribosomal proteins (r-proteins) are transcribed by Pol II (MULLINEUX12, VONWALDEN19). Together, the four mature rRNAs (18S, 5.8S, 28S and 5S) along with 80 ribosomal proteins form the two subunits of a ribosome (FIGUEIREDO19, VONWALDEN19).

Whether glucose has mechanisms to affect ribosome biogenesis i.e., acting as a signalling molecule per se, or indirectly via insulin is currently unknown. Factors involved in the (PIC) of rDNA transcription are thought to receive cues from pathways (mTORC1, MAPK) regulated by hormonal and nutritional signals (FIGUEIREDO19). Furthermore, high glucose treatment promoted ribosome biogenesis while glucose depletion repressed ribosome biogenesis in mice (ZHAI12 – OR WAS IT CELL CULTURES?). Thus, it seems plausible that glucose ingestion before and after RT may benefit ribosome biogenesis. Lastly, there are equivocal observations on insulin’s impact on ribosome biogenesis and MPS. Previous meta-analyses suggest insulin acts in concert with amino acid intake to enhance MPS, and that insulin may reduce MPB when amino acids are scarce (ABDULLA16).

To investigate this further, we recruited 16 young, moderately trained, male and female individuals for heavy unilateral resistance training for two weeks, exercising one leg with glucose and protein, and the other with a placebo and protein ingestion before and after each session. Our main purpose was to investigate the effects of glucose ingestion compared to placebo ingestion before and after five RT sessions on ribosome biogenesis. Changes in ribosome biogenesis were measured by the accumulation of total RNA and expression of the ribosome’s four mature rRNAs and the 47S precursor RNA, along with levels of UBF, c-Myc, and rps6 protein. As a secondary outcome, we were also interested in the effects of glucose compared to placebo ingestion on muscular function and -recovery, measured via unilateral isometric and isokinetic knee extension torque, before, during and after the intervention. We hypothesised that RT would lead to a prominent accumulation of total RNA, rRNA and protein (UBF, c-Myc, rps6) and that RT with glucose and protein would lead to higher accumulation and expression compared to RT with placebo and protein. We also hypothesised that glucose supplementation would enhance muscular function and -recovery compared to placebo.

@@

Total protein was extracted using the Minute Total Protein Extraction Kit for Muscles (Invent Biotechnology), according to the manufacturer’s protocol, optimised for our lab. Wet muscle was freeze-dried for 24hrs and dissected before extraction. The tissue was homogenized with a plastic rod in 80mg protein extraction powder (Invent Biotechnology) and 100ul iced cell lysis buffer (Denaturing Buffer, Invent Biotechnology), and centrifuged at top speed for 1 min. The supernatant was aliquoted into 2x25ul tubes to run samples in duplicates, and total protein concentrations were determined in a 1:10 dilution (Pierce Detergent Compatible Bradford Assay Reagent, Thermo Fisher Scientific). The protein samples were diluted to 20.25 µg µl−1 total protein with lysis buffer and 4X Laemmli sample buffer (Bio-Rad Laboratories AB, Oslo, Norway) containing 2-mercaptoethanol. Twenty microlitres from each protein sample were pooled as a normalisation factor for gel separation. All protein samples were incubated at 95°C and stored at -20°C until further analysis. The protein samples (20.25µg total protein) were separated at 250V for 50min using …. gels (……..), then transferred to PVDF membranes with wet transfer at 300mA for 3hrs. Both gel electrophoresis and protein transfer were performed at 4°C. Following the wet transfer, membranes were stained with a MemCode reversible total protein stain (MANUFACTURER) and then blocked for 1hr at room temp with a blocking buffer of Tris-buffered saline (TBS; 20 mM Tris, 150 mM NaCl) with 5% non-fat dry milk and 0.1% Tween-20. Primary and secondary antibodies were purchased from Santa Cruz Biotechnology (Texas, USA) UBF:(UBF-9, sc-13125), rpS6:(Ribosomal protein S6 C-8, sc-74459), and Thermo Fisher Scientific () c-Myc:(c-Myc 9E10), goat anti-mouse for c-Myc (goat anti-mouse IgG1 (y1) horseradish peroxidase conjugate), and anti-mouse (anti-mouse IgG1 horseradish peroxidase conjugate). Antibodies were diluted in a 5% non-fat dry milk blocking buffer (TBS, 5% non-fat dry milk, 0.1% Tween-20). Membranes were incubated overnight with primary antibodies, and for 1hr with secondary antibodies. Between blocking and primary antibody staining, membranes were washed 1x5min, between primary and secondary staining, and after secondary staining, membranes were washed 3x5 with TBS-Tween (TBS; 20mM Tris, 150mM NaCl, 0.1% Tween). Following the last TBS-T wash, membranes were incubated 1x5min with enhanced chemiluminescent substrate (ECL, SuperSignal West Femto Maximum Sensitivity Substrate, Thermo Fisher Scientific). Membrane blocking, secondary antibody incubation, washing and ECL incubation were performed at room temp, on a tipping board. Primary antibody incubation was performed at 4°C on a tipping board. Chemiluminescence signals were quantified using Image Studio Lite (LI-COR Biotechnology, Lincoln, NE, USA), and total protein content was quantified using ImageJ (RUEDENETAL2017), whereas total protein content was defined as mean grey value of the whole well with between-well values subtracted as background.

@@

## Protein signals

>>>>

Regular resistance exercise training (RT) increases muscle mass and muscle strength, due to the adaptations stimulated by a mechanical strain. Unfortunately, responses to RT are not uniform across the population. Studies have shown RT-induced muscle growth to vary widely between individuals, with as much as 10-15% showing considerably impaired growth (ALVAREZ18, MANN14, THALACKER13). This may be explained by genetics, epigenetics, and/or an unfavourable internal physiological milieu (THALACKER13, HAMMARSTROM20, FIGUEIREDO19?). While tweaking training modalties and -protocols induces different responses in different populations (HAMMARSTROM20, 19?), little is known regarding optimisation of RT to individual phenotypes. Generally, meta-analyses favour a moderate volume of RT over low volume RT regarding muscle growth and -strength (KRIEGER09, RHEA03, SCHOENFELD17), substantiated by recent RCTs (HAMMARSTROM20..). Although Hammarström and colleagues (20) found that a moderate volume led to more prominent gains in muscle mass and -strength, approximately 50% of the participants did not experience true benefits of moderate volume compared to low, with a blunted muscle growth. Due to the variations observed in muscle growth responses despite standardising RT protocols (HAMMARSTROM20; SCHOENFELD17; STEC16?), it is apparent that increasing RT volume alone does not convert low-responders to high-responders. Therefore, other means than RT *per se* seems necessary to circumvent this discrepancy.

Indeed, ingesting nutrients such as protein and creatine supplements effectively optimise RT (CERMAK12, LANHERS15, LANHERS17). However, it remains unknown if other nutritional adjuvants such as glucose can increase the efficacy of RT. This is surprising, as glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major supplier of energy to cells via ATP synthesis (TANAKA18). Energy availability is important to regulate central processes of muscle growth, specifically ribosome biogenesis, a key factor for increased muscle protein synthesis (MPS) (TANAKA18, FIGUEIREDO19, VONWALDEN19?). Interestingly, variations in RT-induced responses have been linked to ribosome biogenesis, where participants responding poorly to RT also showed a blunted ribosome biogenesis (HAMMARSTROM20, STEC16). A blunted ribosome biogenesis negates responses to RT by failure to produce novel ribosomes, thus an inability to increase the skeletal muscle cell’s translational capacity (VONWALDEN19, FIGUEIREDO19). Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into the regulation of ribosome biogenesis.

Ribosome biogenesis is suggested to be regulated by the mammalian target of rapamycin complex 1 (mTORC1) network (KUSNADI15, TANAKA18). Stimulated by growth factors, hormones, mechanical loading and nutrients, mTORC1 governs energy distribution by turning on/off energy-consuming and energy-generating pathways, depending on cellular energy levels (HOPPE09). Ribosome biogenesis consumes up to 80% of the cell’s energy and materials (SCHMIDT04) and is presumably switched on/off based on cellular energy levels (VONWALDEN19). Upon stimulation by mTORC1, ribosome biogenesis is initiated with the assembly of the preinitiation complex (PIC) (VONWALDEN19, FIGUEIREDO19). The PIC consists of the RNA polymerase I complex, the SL-1 complex and the TIF-1A/RRN3 protein (FIGUEIREDO19, VONWALDEN19). Further, the upstream binding factor (UBF) has a vital role in the initiation, binding the PIC to the rDNA promoter, while c-Myc serves as a transcription factor (FIGUEIREDO19, VONWALDEN19). Upon initiation, Pol I transcribes the 47S pre-rRNA, ultimately processed into the 18S-, 5.8S and 28S rRNA (MULLINEUX12, VONWALDEN19). The fourth rRNA, 5S, is transcribed by Pol III, while mRNAs for the ribosomal proteins (r-proteins) are transcribed by Pol II (MULLINEUX12, VONWALDEN19). Together, the four mature rRNAs (18S, 5.8S, 28S and 5S) along with 80 ribosomal proteins form the two subunits of a ribosome (FIGUEIREDO19, VONWALDEN19). The factors involved in the PIC are proposed to receive cues from pathways such as mTORC1 and MAPK, which are regulated by among others hormonal and nutritional signals (FIGUEIREDO19, TANAKA18). High glucose treatment did promote ribosome biogenesis compared to a repressed ribosome biogenesis with low glucose treatment and glucose starvation, in cell cultures (?)(ZHAI12). And so, though it seems apparent that sufficient daily intake of glucose/carbohydrates is necessary for an optimal ribosome biogenesis (ZHAI12, TANAKA18), it is not clear whether there are acute effects on ribosome biogenesis _per se_ with ingesting glucose before and after RT.

Whether glucose has mechanisms to affect ribosome biogenesis i.e., acting as a signalling molecule *per se*, or indirectly via insulin is currently unknown. There are equivocal observations on insulin's impact on ribosome biogenesis and MPS, where previous meta-analyses suggest insulin to act in concert with amino acid intake to enhance MPS (ABDULLA16). Furhter, insulin may reduce skeletal muscle protein breakdown when amino acids are scarce (ABDULLA16). Therefore, to investigate the relationship between glucose, RT and ribosome biogenesis we recruited 16 young, moderately trained, male and female individuals for heavy unilateral resistance training for two weeks. They exercised one leg with glucose and protein, and the other leg with placebo and protein ingestion before and after each RT-session. Our main purpose was to investigate whether there are acute effects of glucose- compared to placebo ingestion before and after RT. To assess this purpose, we measured changes in ribosome biogenesis as accumulation of total RNA, expression of the four mature rRNA's (18S, 5.8S, 28S, 5S), 47S pre-rRNA, and levels of proteins associated with transcription initiation (c-Myc, UBF, rpS6?). As secondary outcomes, we measured the developement in skeletal muscule performance during the intervention, and skeletal muscle recovery after the final RT session, as temporal changes in maximal isometric and isokinetic torque. We hypothesised that RT would indeed lead to a prominent accumulation of total RNA, increased expression of rRNA and associated proteins. Further, we hypothesised that RT with glucose and protein would induce an enhanced response compared to placebo, in markers of ribosome biogenesis, as well as muscular performance and -recovery.

@@

Total protein was extracted using the Minute Total Protein Extraction Kit for Muscles (Invent Biotechnology), according to the manufacturer’s protocol, optimised for our lab. Wet muscle was freeze-dried for 24hrs and dissected before extraction. The tissue was homogenized with a plastic rod in 80mg protein extraction powder (Invent Biotechnology) and 100ul iced cell lysis buffer (Denaturing Buffer, Invent Biotechnology), and centrifuged at top speed for 1 min. The supernatant was aliquoted into 2x25ul tubes to run samples in duplicates, and total protein concentrations were determined in a 1:10 dilution (Pierce Detergent Compatible Bradford Assay Reagent, Thermo Fisher Scientific, Oslo, Norway). The protein samples were diluted to 20.25 µg µl−1 total protein with lysis buffer and 4X Laemmli sample buffer (Bio-Rad Laboratories AB, Oslo, Norway) containing 2-mercaptoethanol. Twenty microlitres from each protein sample were pooled as a normalisation factor for gel separation. All protein samples were incubated at 95°C and stored at -20°C until further analysis. The protein samples (20.25µg total protein) were separated at 250V for 50min using 4-20% Tris-Glycine gels (Thermo Fisher Scientific), then transferred to PVDF membranes with wet transfer at 300mA for 3hrs. Both gel electrophoresis and protein transfer were performed at 4°C. Following the wet transfer, membranes were stained with a MemCode reversible total protein stain (MANUFACTURER) and then blocked for 1hr at room temp with a blocking buffer of Tris-buffered saline (TBS; 20 mM Tris, 150 mM NaCl) with 5% non-fat dry milk and 0.1% Tween-20. Primary and secondary antibodies were purchased from Santa Cruz Biotechnology (Texas, USA) UBF:(UBF-9, sc-13125), rpS6:(Ribosomal protein S6 C-8, sc-74459), and Thermo Fisher Scientific () c-Myc:(c-Myc 9E10), goat anti-mouse for c-Myc (goat anti-mouse IgG1 (y1) horseradish peroxidase conjugate), and anti-mouse (anti-mouse IgG1 horseradish peroxidase conjugate). Antibodies were diluted in a 5% non-fat dry milk blocking buffer (TBS, 5% non-fat dry milk, 0.1% Tween-20). Membranes were incubated overnight with primary antibodies, and for 1hr with secondary antibodies. Between blocking and primary antibody staining, membranes were washed 1x5min, between primary and secondary staining, and after secondary staining, membranes were washed 3x5 with TBS-Tween (TBS; 20mM Tris, 150mM NaCl, 0.1% Tween). Following the last TBS-T wash, membranes were incubated 1x5min with enhanced chemiluminescent substrate (ECL, SuperSignal West Femto Maximum Sensitivity Substrate, Thermo Fisher Scientific). Membrane blocking, secondary antibody incubation, washing and ECL incubation were performed at room temp, on a tipping board. Primary antibody incubation was performed at 4°C on a tipping board. Chemiluminescence signals were quantified using Image Studio Lite (LI-COR Biotechnology, Lincoln, NE, USA), and total protein content was quantified using ImageJ (RUEDENETAL2017), whereas total protein content was defined as mean grey value of the whole well with between-well values subtracted as background.

@@

### Protein signals
